Fast Sorting on Modern Computers

### 2005-08-02 docs/tech_report/elementary.tex
added testing comments
merged david's elementary changes

\code{Selection sort code}{selectsort.c}
\subsection{Testing}
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

Selection sort uses approximately $N^2/2$ comparisons and exactly $N$ exchanges.
As a result, selection sort can be very useful for sorts involving very large
records, and small keys. In this case, the cost of swapping keys dominates the
cost of comparing them. Conversely, for large keys with small records, the cost
of selection sort is far higher than for other simple sorts.

@@

\label{selection branches logn}
The branch prediction performance is not expected to be bad. Flow control
predictions are very straightforward, and should result in very few misses.
Comparison predictions, however, are very numerous. Each array traversal has an
average of $N/2$ comparisons, and approximately $log_2(N/2)$ of them should be
taken, while the rest are not-taken. The taken predictions will be misses,
while the rest should be hits, so it is expected that there will be $logN$
misses per key.

This number is not immediately obvious. Consider a single step of the selection
sort. The first key to be sorted is, on average, larger than half the keys and
smaller than the other half, so it will have to travel across half the array to
find a key smaller than it, on average. Once it finds a smaller key, this will
be larger than a quarter of the keys in the array, and so it will need to travel
half of the remaining array, on average, before it finds it. The search will
continue like this until the entire array has been covered. Comparing two keys
and discovering that the new key is not smaller than the current smallest is the
norm; every time a smaller key is found a branch misprediction will occur (if
two or more keys in a row are successively smaller, this may not be the case).
The length of the array to be considered in $N/2$ on average, and an average
total of $Nlog_2(N/2)$ branch misses will occur. This assumes randomly sorted
keys.

%TODO look at knuths number here
TODO Knuth has a formula which I will look up.

@@

The branch predictive performance is also expected to be similar. While
selection sort has $O(NlogN)$ misses, each of bubblesort's comparisons has
the potential to be a miss, and not just in the pathological case. However, as
the sort continues and the array becomes more sorted, the number of misses
should decrease. The number of flow control misses should be very low.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse, reducing its number of misses. Improved
shakersort is placed roughly between insertion sort and shakersort, showing that
ending early is helpful in its case. This does not appear to be the case with
improved bubblesort though, which performs identically to bubblesort.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key.  Branch prediction performance is not surprising.
Each sort has a similar number of branches, with insertion sort having half as
many as selection and bubblesort, and shakersort being half-way between the
two. The number of misses for bubblesort and shakersort are huge.  As the data
sets gets large, there are nearly 10,000 branch misses involved in sorting a
single key. The improved bubblesort had exactly the same number of misses and
branches. The improved shakersort has a lot less branches, but the same number
of misses. This is because skipping over sorted keys is predictable from the
first branch miss.

Selection sort, meanwhile, has the same number of branches as bubblesort, but
only $O(logN)$ misses per key. The most interesting result, though, is insertion
sort. This has exactly one miss per key. Although this result was not predicted
in advance, on reflection this behaviour is obvious: every iteration involves
moving the key left. The only time it will not move left is when its being put
into place, and this only happens once for every key. This shows that it is
possible to completely eliminate even comparative branch misses. It also
establishes insertion sort as the best of the simple sorts, and it is used later
in quicksort and mergesort as a result of this. \label{insertion is
predictable}

Another interesting result comes from a comparison between shakersort and
selection sort. The number of cycles to sort an array (shown in Figure
\ref{elementary cycles}) is nearly twice as high on shakersort as it is for
selection sort. Shakersort has a significantly lower instruction count and cache
miss count than selection sort, but has several orders of magnitude worse branch
misses. This shows how important it is to consider branch misses in an
algorithm, especially comparative ones. It also establishes that exchange-based
algorithms, such as bubblesort and shakersort, will perform considerably worse
than algorithms which have a correlation between successive branches, as in
insertion and selection sorts.

>>>>

\code{Selection sort code}{selectsort.c}
\subsection{Testing}
%TESTING
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

Selection sort uses approximately $N^2/2$ comparisons and exactly $N$ exchanges.
As a result, selection sort can be very useful for sorts involving very large
records, and small keys. In this case, the cost of swapping keys dominates the
cost of comparing them. Conversely, for large keys with small records, the cost
of selection sort may be higher than for other simple sorts.

@@

\label{selection branches logn}
The branch prediction performance is not expected to be bad. Flow
control predictions are very straightforward, and should result in
very few misses.  Comparison predictions, however, are very
numerous. Each traversal of the array has an average of $N/2$
comparisons, as we search for the smallest key in the unsorted part of
the array. During the search, the algorithm compares each key with the
smallest key seen so far. In a random array, we would expect that the
candidate for the smallest key (the left-to-right minimum) would
change frequently towards the start of the array, making the
comparison branch rather unpredictable. However, as the traversal
continues, the left-to-right minimum will become smaller, and thus
more difficult to displace. It will become very unusual to find a
smaller key, and thus the branch will almost always resolve in the
same direction.  Therefore, the comparison branch will quickly become
very predictable.

Knuth \cite{Knuth1} analyses the number of changes to the
left-to-right minimum when searching for the minimum element of an
array. He shows that for $N$ elements, the number of changes is $H_N -
1$, where $H_N$ is the harmonic series:

\displaymath{H_N = 1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{N} =  \sum_{k=1}^{N} \frac{1}{k},  \hspace{0.4cm}  N \geq 1}

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log(N)$. Thus, in the average case, we can expect that the comparison
branch will mostly go in the same direction, making the branch highly
predictable.

Knuth also presents the number of changes to the left-to-right minimum
in the best case (0 changes) and worst cases ($N-1$ changes).  In the
worst case, the array is sorted in reverse order, and so the
left-to-right minimum changes on every element. Interestingly, the
worst cases for the number of changes is not the worst case for branch
mispredictions. If the minimum changes for every element, then the
comparison branch will always resolve in the same direction, and thus
be almost perfectly predictable.  From the point of view of branch
prediction, frequent changes (preferably with no simple pattern) are
the main indicator of poor performance, rather than the absolute
number of times that the branch resolves in each direction. The worst
case for selection sort would probably be an array which is partially
sorted in reverse order --- the left-to-right minimum would change
frequently, but not predictably so.

@@

The branch predictive performance is also expected to be similar. While
selection sort has $O(NH_N)$ mispredictions, each of bubblesort's comparisons
has the potential to be a misprediction, and not just in the pathological case.
However, as the sort continues and the array becomes more sorted, the number of
mispredictions should decrease. The number of flow control mispredictions should
be very low.

@@

As expected, selection sort has the worst cache performance. This is due to the
fact that it considers the entire unsorted segment of the array for each key to
be put in place. Bubblesort, with similar behaviour, is only slightly better.
Shakersort has a small amount of temporal reuse, reducing its number of misses.
Improved shakersort is placed roughly between insertion sort and shakersort,
showing that ending early is helpful in its case. This does not appear to be the
case with improved bubblesort though, which performs identically to bubblesort.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key. Due to huge variations in predictability, we split
the numbers of branch mispredictions into two charts. Selection sort causes
around $H_N$ mispredictions per key, which is remarkably small. The comparison
branch in the inner loop is highly predictable on random data, because a good
candidate minimum is usually found quickly.

Selection sort, meanwhile, has the same number of branches as bubblesort, but
only $O(H_N)$ misses per key. The most interesting result, though, is insertion
sort. This has exactly one miss per key. Although this result was not predicted
in advance, on reflection this behaviour is obvious: every iteration involves
moving the key left. The only time it will not move left is when its being put
into place, and this only happens once for every key. This shows that it is
possible to completely eliminate even comparative branch misses. It also
establishes insertion sort as the best of the simple sorts, and it is used later
in quicksort and mergesort as a result of this. \label{insertion is predictable}

The number of misses for bubblesort and shakersort are so huge that we had to
plot them on a separate chart. Nearly one in four comparisons are misses. As the
data sets gets large, there are nearly 10,000 branch misses involved in sorting
a single key. The improved bubblesort had exactly the same number of misses and
branches. The improved shakersort has a lot less branches, but the same number
of misses. This is because skipping over sorted keys is predictable from the
first branch miss.

The branch behaviour of bubble sort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubble sort finds the smallest (or largest) key in the
unsorted part of the array and moves it to start (end) of the array.
Thus, we would expect that bubbble sort's branch prediction behaviour
would be similar to that of selection sort.

However, in the proces of moving the smallest element into position,
bubble sort also rearranges other array elements. Selection sort
sweeps across the array and always keeps track of the left-to-right
minimum up to the current point. However, bubble sort actually moves
this minimum, exchanging pairwise with each element it encounters
which is larger. Thus, even the unsorted part of the array becomes
partially sorted. We might expect that bubble sort's comparison branch
for a partially sorted array might be more predictable than for a
random one, but it is not. With a random array, the smaller elements
are equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubble sort, the last place that we look. Thus,
we change the candidate minimum much more often than if we were
searching a randomly ordered array.
